\chapter{Abstract}
\vspace*{-0.75cm}
Gravitational-wave (GW) astronomy has ushered in a new era of scientific exploration, providing us with unprecedented insights into compact binary coalescing sources, such as binary black holes and neutron star binaries. GW detectors like Advanced LIGO and Virgo have observed almost a hundred merging binaries, but their origins remain unsolved. The presence of significant orbital eccentricity or orbital precession are signatures for binaries formed in dense environments or with multi-body interactions during their evolution. If such binaries are detected they would indicate the presence of a dynamical formation channel. Detection of these systems could also provide insights into long standing uncertain astrophysical and physical processes such as common envelope evolution, supernovae natal kicks and the dynamics of dense environments. Typically, searches are performed for binaries with quasi-circular orbits, spins aligned to the orbital angular momentum and capture only the dominant mode of the gravitational radiation. Relaxing any one of these search assumptions requires up to $100\times$ larger computational power than an equivalent non-eccentric, non-precessing search. In this thesis we push the boundaries of existing search pipelines to search for novel binaries. We tackle this problem in three different avenues. First, by extending current search methods to look for eccentric systems -- we performed the first search for spinning eccentric neutron star binaries in the public data of Advanced LIGO and Virgo observatories. Using our search results we put state-of-art observational constraints on various astrophysical models and predict eccentric observations for the future observatories. Second, we investigate how many precessing sources may have been missed by existing searches, and identify regions of parameter space crucial for targeted precessing searches. Finally, we address a common issue for searching either of the two novel types of binaries -- increased computational costs. We demonstrate a new matched filtering technique that may save up to $10\times$ the computational costs without losing sensitivity and can be applied to any modeled search schemes.    


%Eccentric binaries are characterized by elliptical orbits, deviating from the more common circular orbits, while precessing binaries experience a gradual rotation of their orbital plane due to the objects' individual spins. The search for such systems poses significant computational challenges, as the increased complexity of their waveforms necessitates exhaustive exploration of a larger parameter space.

%To overcome these challenges, researchers have been developing new search methods that leverage innovative techniques and strategies. One promising avenue is the adoption of hierarchical search strategies. These strategies involve breaking down the search process into multiple stages, allowing for a systematic exploration of parameter space while optimizing computational resources. In a hierarchical search, the initial stage focuses on rapidly identifying promising candidates by employing computationally inexpensive algorithms that are less sensitive to the intricate details of eccentricity and precession. Subsequent stages then refine the analysis by employing more computationally intensive algorithms to extract the subtle signatures associated with these specific binary configurations.

%By implementing hierarchical search strategies, researchers aim to improve the detection efficiency and sensitivity of gravitational wave detectors, enabling the discovery of previously elusive sources. These new methods not only have the potential to uncover a wealth of knowledge about the astrophysical processes governing eccentric and precessing binaries but also contribute to our understanding of general relativity and the formation and evolution of compact objects in the universe.

%In conclusion, the development of new search methods for detecting gravitational wave signals from compact binary coalescing sources, particularly eccentric or precessing binaries, is an active and exciting area of research. Hierarchical search strategies offer a promising approach to overcome the computational challenges associated with these systems, providing a pathway towards the discovery and characterization of a wider range of gravitational wave sources. By pushing the boundaries of detection capabilities, these advancements will undoubtedly deepen our understanding of the cosmos and bring us closer to unlocking the secrets of the universe.

%in order to maximise the detection rate, or have limited computational power
\clearpage
\thispagestyle{plain}
\textbf{Keywords:} gravitational waves, compact binary mergers, gravitational-wave search algorithms
\thispagestyle{plain}
