\chapter{First ever search for eccentric NSBH and the most sensitive search for BNS mergers}

This chapter briefly summarizes the search performed in the paper []. We had target a previously unexplored region of the parameter space for search of eccentric binary systems. We did not find any new significant events, but explore upper limits for four different astrophysical models which have a wide range of astrophysical and physical implications.  

\section{Introduction}

The origins of compact binary systems is still unknown: their formation and evolution mechanism remains elusive. Various formation models exist in the literature, which are broadly categorized in two categories -- isolated binary channels and dynamical formation scenarios. In section [], we briefly discussed the various possible mechanisms for the two broad channels. Each of the model has a unique prediction of the distribution of binary parameters -- chirp mass $\mathcal{M}_c$, orbital eccentricity (at a given reference frequency) and spins of the compact objects. 

The primary difference between the two channels is their predictions about the component spins and the orbital eccentricity. Isolated formation channels predict component spins to be nearly aligned with the orbital angular momentum, whereas, dynamical scenarios may have asymmetric spins. Another robust signature of a binary's evolutionary history is the orbital eccentricity. Compact binaries from any channel may have very high eccentricities ($e \sim 1$) because of the asymmetric kicks from the supernovae explosion of the stellar objects []. However, the subsequent evolution of a binary might be different depending on the environment. Isolated binaries radiate energy via the GW emission only. And, as discussed in section (), eccentricity quickly vanishes with the evolution of orbital frequency and the system circularizes when the merger enters the sensitive bands of the current GW observatories (i.e 10 Hz). However, in dense environments angular exchanges with a tertiary object can induce eccentricity via the Lidov-Kozai (LK) or (Zeipel-Lidov-Kozai) mechanism as explained in section []. These mechanisms retain high eccentricities $e_{10} \geq 0.01$ at the observable frequencies. Thus, observation of an eccentric system would clearly indicate the presence of a dynamical channel.

The various formation models also come with a prediction for the local merger rates.      


Later, in section [], we discussed the most recent gravitational wave catalogs. The current observations of LIGO and VIRGO events are broadly consistent with isolated formation channels and are not sufficient enough to distinguish the two broad channels.  Distinguishing between isolated or dynamical channels will help us improve our understanding of the standard formation scenarios. In case if we detect special events, we might be able to rule out certain models. For example stellar evolution does not predict any black-holes above $65 M_{\odot}$ and the detection of the heaviest event GW190521 with component masses $142^{+28}_{-16} M_{\odot}$ and $85^{+21}_{-14} M_{\odot}$. 

\section{Search setup}